Comparing Models of Delta-Notch Signalling

### 2025-04-11 paper/paper.tex
Point to new vis11 figure paper.tex
paper.tex
Add a note to Table 1 description; Update KFE; Update Theoretical Justification

\begin{table}[!htp]
\centering
\begin{tabular}{|m{8em}|m{5em}|m{20em}|} 
 \hline
 Reaction & Rate & Description \\ 
 \hline
 $N_{A} + D_{B} \rightarrow I_{A}$ & 
 $k_{T} N_{A}D_{B}$ &
 A Notch receptor from cell $A$ binds to a delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
 \hline
 $N_{A} \rightarrow \emptyset$ & 
 $\gamma N_{A}$ & 
 A Notch receptor from cell $A$ decays. \\
 \hline
 $D_{A} \rightarrow \emptyset$ & 
 $\gamma D_{A}$ & 
 A Delta receptor from cell $A$ decays. \\
 \hline
 $I_{A} \rightarrow \emptyset$ &
 $\gamma_{I} I_{A}$ &
 A NICD molecule from cell $A$ decays.  \\
 \hline
 $\emptyset \rightarrow N_{A}$ & 
 $H^{+}(I_{A})$ &
 A Notch receptor in cell $A$ is produced \\
 \hline
 $\emptyset \rightarrow D_{A}$ &
 $H^{-}(I_{A})$ &
 A Delta ligand in cell $A$ is produced \\
 \hline
\end{tabular}
\caption{
  A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
}
\label{tb:reactions}
\end{table}

@@

\begin{figure}[!htp]
  \includegraphics[width=\textwidth]{img/vis11.png}
  \caption{A plot of the KDE approximation of the distribution of differentiation times for the agent-based model and the stochastic ODE model with different levels of noise. All simulations were done on a two-cell domain with periodic boundary conditions. These simulations show that higher values of the noise parameter significantly decrease the average time it takes for the two cells to adopt different fates. Notably, a noise parameter of $2\%$ causes the stochastic ODE model to align most closely with the results of agent-based simulations. }
  \label{fig:noise-estimate}
\end{figure}

@@

% TODO: Revise this section to incorporate Eric's feedback.

\subsection*{Derivation of Kolmogorov Forward Equation}

Let us define $P(n,d,i,t)$ to be the probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules at time $t$. The Kolmogorov forward equation describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:

\begin{align*}
P(n,d,i,t+\Delta t) &= P(n,d,i,t) \\
&+ P(n-1,d,i,t) \cdot H^+(i) \Delta t \quad \text{(Notch production)} \\
&+ P(n,d-1,i,t) \cdot H^-(i) \Delta t \quad \text{(Delta production)} \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1)\Delta t \quad \text{(Notch decay)} \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1)\Delta t \quad \text{(Delta decay)} \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1)\Delta t \quad \text{(NICD decay)} \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext}\Delta t \quad \text{(Trans-activation)} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]\Delta t
\end{align*}

Rearranging to find the time derivative:

\begin{align*}
\frac{P(n,d,i,t+\Delta t) - P(n,d,i,t)}{\Delta t} &= P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\end{align*}

Taking the limit as $\Delta t \rightarrow 0$:

\begin{align*}
\frac{\partial P(n,d,i,t)}{\partial t} &= P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\end{align*}

\subsection*{Matrix Formulation}
The Kolmogorov forward equation can be expressed in matrix form as:
\[
\frac{d\vec{P}(t)}{dt} = \mathbf{A} \vec{P}(t)
\]

where $\vec{P}(t)$ is a column vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix. \\

To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$. For a system with maximum values $n_{max}$, $d_{max}$, and $i_{max}$, we define:
\[
k(n,d,i) = n \cdot (d_{max}+1) \cdot (i_{max}+1) + d \cdot (i_{max}+1) + i
\]
for $0 \leq n \leq n_{max}, \; 0 \leq d \leq d_{max}, \; 0 \leq i \leq i_{max}$.

@@

\begin{align*}
A_{k',k} = 
\begin{cases}
H^+(i) & \text{if } (n',d',i') = (n+1,d,i) \quad \text{(Notch production)} \\
H^-(i) & \text{if } (n',d',i') = (n,d+1,i) \quad \text{(Delta production)} \\
\gamma(n) & \text{if } (n',d',i') = (n-1,d,i) \quad \text{(Notch decay)} \\
\gamma(d) & \text{if } (n',d',i') = (n,d-1,i) \quad \text{(Delta decay)} \\
\gamma_I(i) & \text{if } (n',d',i') = (n,d,i-1) \quad \text{(NICD decay)} \\
k_T(n)D_{ext} & \text{if } (n',d',i') = (n-1,d,i+1) \quad \text{(Trans-activation)} \\
-\sum_{k'' \neq k} A_{k'',k} & \text{if } k' = k \quad \text{(Diagonal terms)}
\end{cases}
\end{align*}

Note that $A_{k',k}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:

\[
A_{k,k} = -[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}]
\]

where $(n,d,i)$ is the state corresponding to index $k$.\\

@@

To illustrate, let's consider a minimal system with $n_{max} = d_{max} = i_{max} = 1$. This gives us 8 possible states:
$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$

@@

\begin{align*}
\mathbf{A} = 
\begin{pmatrix}
-\alpha_{000} & \gamma_I & \gamma & 0 & \gamma & 0 & 0 & 0 \\
H^+(0) & -\alpha_{001} & 0 & \gamma & 0 & \gamma & 0 & 0 \\
H^-(0) & 0 & -\alpha_{010} & \gamma_I & 0 & 0 & \gamma & 0 \\
0 & H^-(0) & H^+(0) & -\alpha_{011} & 0 & 0 & 0 & \gamma \\
H^+(0) & 0 & 0 & 0 & -\alpha_{100} & \gamma_I & \gamma & 0 \\
0 & H^+(0) & 0 & 0 & H^+(1) & -\alpha_{101} & 0 & \gamma \\
0 & 0 & H^+(0) & 0 & H^-(1) & 0 & -\alpha_{110} & \gamma_I \\
0 & 0 & 0 & H^+(0) & 0 & H^-(1) & H^+(1) & -\alpha_{111}
\end{pmatrix}
\end{align*}

where:
\[
\alpha_{ndi} = H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} + k_T d N_{ext}
\]

Additionally, we need to add the terms for trans-activation, which would connect states like $(1,0,0)$ to $(0,0,1)$ with rate $k_T \cdot 1 \cdot D_{ext}$.\\

For a fully labelled matrix with all transitions explicitly shown, the matrix for $n_{max} = d_{max} = i_{max} = 1$ would be:

\[
\mathbf{A} = 
\begin{bmatrix}
A_{11} & A_{12} & A_{13} & A_{14} & A_{15} & A_{16} & A_{17} & A_{18} \\
B_{11} & B_{12} & B_{13} & B_{14} & B_{15} & B_{16} & B_{17} & B_{18} \\
C_{11} & C_{12} & C_{13} & C_{14} & C_{15} & C_{16} & C_{17} & C_{18} \\
D_{11} & D_{12} & D_{13} & D_{14} & D_{15} & D_{16} & D_{17} & D_{18} \\
E_{11} & E_{12} & E_{13} & E_{14} & E_{15} & E_{16} & E_{17} & E_{18} \\
F_{11} & F_{12} & F_{13} & F_{14} & F_{15} & F_{16} & F_{17} & F_{18} \\
G_{11} & G_{12} & G_{13} & G_{14} & G_{15} & G_{16} & G_{17} & G_{18} \\
H_{11} & H_{12} & H_{13} & H_{14} & H_{15} & H_{16} & H_{17} & H_{18}
\end{bmatrix}
\]

where

\begin{tiny}
\begin{align*}
A_{11} &= -(H^+(0) + H^-(0)),  & A_{12} &= \gamma_I,  \\
A_{13} &= \gamma,  & A_{14} &= 0,  \\
A_{15} &= \gamma,  & A_{16} &= 0,  \\
A_{17} &= 0,  & A_{18} &= 0,  \\[5pt]
B_{11} &= H^+(0),  & B_{12} &= -(H^+(1) + H^-(1) + \gamma_I),  \\
B_{13} &= 0,  & B_{14} &= \gamma,  \\
B_{15} &= k_T D_{ext},  & B_{16} &= \gamma,  \\
B_{17} &= 0,  & B_{18} &= 0,  \\[5pt]
C_{11} &= H^-(0),  & C_{12} &= 0,  \\
C_{13} &= -(H^+(0) + H^-(0) + \gamma + k_T N_{ext}),  & C_{14} &= \gamma_I,  \\
C_{15} &= 0,  & C_{16} &= 0,  \\
C_{17} &= \gamma,  & C_{18} &= 0,  \\[5pt]
D_{11} &= 0,  & D_{12} &= H^-(1),  \\
D_{13} &= H^+(0),  & D_{14} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T N_{ext}),  \\
D_{15} &= 0,  & D_{16} &= k_T D_{ext},  \\
D_{17} &= 0,  & D_{18} &= \gamma,  \\[5pt]
E_{11} &= H^+(0),  & E_{12} &= 0,  \\
E_{13} &= 0,  & E_{14} &= 0,  \\
E_{15} &= -(H^+(0) + H^-(0) + \gamma + k_T D_{ext}),  & E_{16} &= \gamma_I,  \\
E_{17} &= \gamma,  & E_{18} &= 0,  \\[5pt]
F_{11} &= 0,  & F_{12} &= H^+(1),  \\
F_{13} &= 0,  & F_{14} &= 0,  \\
F_{15} &= H^+(0),  & F_{16} &= -(H^+(1) + H^-(1) + \gamma + \gamma_I + k_T D_{ext}),  \\
F_{17} &= 0,  & F_{18} &= \gamma,  \\[5pt]
G_{11} &= 0,  & G_{12} &= 0,  \\
G_{13} &= H^+(0),  & G_{14} &= 0,  \\
G_{15} &= H^-(0),  & G_{16} &= 0,  \\
G_{17} &= -(H^+(0) + H^-(0) + \gamma + \gamma + k_T D_{ext} + k_T N_{ext}),  & G_{18} &= \gamma_I,  \\[5pt]
H_{11} &= 0,  & H_{12} &= 0,  \\
H_{13} &= 0,  & H_{14} &= H^+(1),  \\
H_{15} &= 0,  & H_{16} &= H^-(1),  \\
H_{17} &= H^+(0),  & H_{18} &= -(H^+(1) + H^-(1) + \gamma + \gamma + \gamma_I + k_T D_{ext} + k_T N_{ext}).
\end{align*}
\end{tiny}

Each row and column corresponds to a state in the order:
$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1)$

The matrix element $A_{j,i}$ represents the transition rate from state $i$ to state $j$. The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.

@@

% TODO: Revise this section to incorporate Eric's feedback.

In this section, we need to show that in expectation, the model above converges to the following system of ODEs (adapted from Boareto et al., 2015):

$$
\begin{aligned}
  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
\end{aligned}
$$

@@

Let us define:
\begin{itemize}
  \item $\langle N \rangle = \sum_{n,d,i} n \cdot P(n,d,i,t)$ = expected number of Notch receptors
  \item $\langle D \rangle = \sum_{n,d,i} d \cdot P(n,d,i,t)$ = expected number of Delta ligands
  \item $\langle I \rangle = \sum_{n,d,i} i \cdot P(n,d,i,t)$ = expected number of NICD molecules
\end{itemize}

\subsection*{Deriving ODEs for Expected Values}

\textbf{Now we derive the ODE for the expected value of Notch:}

\begin{align*}
\frac{d\langle N \rangle}{dt} &= \frac{d}{dt}\sum_{n,d,i} n \cdot P(n,d,i,t) \\
&= \sum_{n,d,i} n \cdot \frac{\partial P(n,d,i,t)}{\partial t}
\end{align*}

Substituting the Kolmogorov forward equation:

\begin{align*}
\frac{d\langle N \rangle}{dt} &= \sum_{n,d,i} n \cdot \Big[ P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}] \Big]
\end{align*}

Now we use index shifting to simplify.\\~\\

\textbf{(A) Notch Production} \\
Let's substitute $n' = n-1$ (so $n = n'+1$):
\begin{align*}
\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n',d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n',d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n',d,i} P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n,d,i} n \cdot P(n,d,i,t) \cdot H^+(i) + \sum_{n,d,i} P(n,d,i,t) \cdot H^+(i) \\
\end{align*}

The first term is $\langle N \cdot H^+(I) \rangle$. If we assume that $H^+(I)$ is approximately $H^+(\langle I \rangle)$, then this term becomes $\langle N \rangle \cdot H^+(\langle I \rangle)$. The second term is $\langle H^+(I) \rangle \approx H^+(\langle I \rangle)$.

So the contribution of this term is:
\[
\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) \approx \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle)
\]

Since we are indexing from 0, and $n-1$ could be negative, the sum should start from $n=1$:

\begin{small}
\begin{align*}
\sum_{n \geq 1,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n' \geq 0,d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n' \geq 0,d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n' \geq 0,d,i} P(n',d,i,t) \cdot H^+(i) \\
&= \langle N \rangle \cdot H^+(\langle I \rangle) + H^+(\langle I \rangle) \\
&= H^+(\langle I \rangle) \cdot (1 + \langle N \rangle)
\end{align*}
\end{small}

\textbf{(B) Delta Production}\\
This term does not affect $N$, so the prefactor $n$ remains with the original probability:
\begin{align*}
\sum_{n,d,i} n \cdot P(n,d-1,i,t) \cdot H^-(i) &= \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot H^-(i) \\
&= \langle N \cdot H^-(I) \rangle \\
&\approx \langle N \rangle \cdot H^-(\langle I \rangle)
\end{align*}

\textbf{(C) Notch Decay}
\begin{align*}
\sum_{n,d,i} n \cdot P(n+1,d,i,t) \cdot \gamma(n+1) &= \sum_{n',d,i} (n'-1) \cdot P(n',d,i,t) \cdot \gamma n' \\
&= \gamma \sum_{n',d,i} (n'-1)n' \cdot P(n',d,i,t) \\
&= \gamma \sum_{n',d,i} n'^2 \cdot P(n',d,i,t) - \gamma \sum_{n',d,i} n' \cdot P(n',d,i,t) \\
&= \gamma(\langle N^2 \rangle - \langle N \rangle)
\end{align*}

If we assume that the variance of $N$ is small compared to its mean (which is reasonable for high molecule counts), then $\langle N^2 \rangle \approx \langle N \rangle^2$. So:
\begin{align*}
\gamma(\langle N^2 \rangle - \langle N \rangle) &\approx \gamma(\langle N \rangle^2 - \langle N \rangle) \\
&= \gamma \langle N \rangle (\langle N \rangle - 1)
\end{align*}

>>>>

\begin{table}[!htp]
\centering
\begin{tabular}{|m{8em}|m{5em}|m{20em}|} 
 \hline
 Reaction & Rate & Description \\ 
 \hline
 $N_{A} + D_{B} \rightarrow I_{A}$ & 
 $k_{T} N_{A}D_{B}$ &
 A Notch receptor from cell $A$ binds to a delta ligand from cell $B$, triggering the release of NICD in cell $A$. \\
 \hline
 $N_{A} \rightarrow \emptyset$ & 
 $\gamma N_{A}$ & 
 A Notch receptor from cell $A$ decays. \\
 \hline
 $D_{A} \rightarrow \emptyset$ & 
 $\gamma D_{A}$ & 
 A Delta receptor from cell $A$ decays. \\
 \hline
 $I_{A} \rightarrow \emptyset$ &
 $\gamma_{I} I_{A}$ &
 A NICD molecule from cell $A$ decays.  \\
 \hline
 $\emptyset \rightarrow N_{A}$ & 
 $H^{+}(I_{A})$ &
 A Notch receptor in cell $A$ is produced \\
 \hline
 $\emptyset \rightarrow D_{A}$ &
 $H^{-}(I_{A})$ &
 A Delta ligand in cell $A$ is produced \\
 \hline
\end{tabular}
\caption{
  A list of possible reactions and their rates in a cell $A$ with neighbour $B$.
  Since there are $6$ possible reactions per cell, a system with $n$ cells will have $6n$ possible reactions.
  \textbf{Note:} In the Reaction column, $N_A$, $D_B$, and $I_A$ represent one Notch, Delta, and NICD, respectively. In the Rate column, $N_A$, $D_B$, and $I_A$ refer to the number of Notch molecules in cell A, Delta in cell B, and NICD in cell A, respectively.
}
\label{tb:reactions}
\end{table}

@@

\begin{figure}[!htp]
  \includegraphics[width=\textwidth]{img/vis11_V2.png}
  \caption{A plot of the KDE approximation of the distribution of differentiation times for the agent-based model and the stochastic ODE model with different levels of noise. All simulations were done on a two-cell domain with periodic boundary conditions. These simulations show that higher values of the noise parameter significantly decrease the average time it takes for the two cells to adopt different fates. Notably, a noise parameter of $2\%$ causes the stochastic ODE model to align most closely with the results of agent-based simulations. }
  \label{fig:noise-estimate}
\end{figure}

@@

\subsection*{Derivation of Kolmogorov Forward Equation}
For simplicity, we consider the dynamics of a single cell (cell $A$) in the following derivations.

Let us define:
\begin{itemize}
    \item $P(n,d,i,t)$: the probability of having $n$ Notch receptors, $d$ Delta ligands, and $i$ NICD molecules in cell $A$ at time $t$.
    \item $D_{ext}$: the average number of Delta molecules in the neighbourhood of cell $A$.
\end{itemize}

The Kolmogorov forward equation describes how $P(n,d,i,t)$ changes over a small time step $\Delta t$:
\begin{align*}
P(n,d,i,t+\Delta t) &= P(n,d,i,t) \\
&+ P(n-1,d,i,t) \cdot H^+(i) \Delta t \quad \text{(Notch production)} \\
&+ P(n,d-1,i,t) \cdot H^-(i) \Delta t \quad \text{(Delta production)} \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1)\Delta t \quad \text{(Notch decay)} \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1)\Delta t \quad \text{(Delta decay)} \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1)\Delta t \quad \text{(NICD decay)} \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext}\Delta t \quad \text{(Trans-activation)} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]\Delta t
\end{align*}

Rearranging to find the time derivative and taking the limit as $\Delta t \rightarrow 0$:
\begin{align*}
\frac{\partial P(n,d,i,t)}{\partial t} &= P(n-1,d,i,t) \cdot H^+(i) \\
&+ P(n,d-1,i,t) \cdot H^-(i) \\
&+ P(n+1,d,i,t) \cdot \gamma(n+1) \\
&+ P(n,d+1,i,t) \cdot \gamma(d+1) \\
&+ P(n,d,i+1,t) \cdot \gamma_I(i+1) \\
&+ P(n+1,d,i-1,t) \cdot k_T(n+1)D_{ext} \\
&- P(n,d,i,t) \cdot [H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\end{align*}

\subsection*{Matrix Formulation}
The Kolmogorov forward equation can be expressed in matrix form as:
\[
\frac{d\vec{P}(t)}{dt} = \vec{P}(t)\mathbf{A}
\]
where $\vec{P}(t)$ is a row vector containing the probabilities for all possible states, and $\mathbf{A}$ is the transition rate matrix with $A_{i,j}$ giving the transition rate from state $i$ to $j$. The diagonal elements $A_{i,i}$ contain the negative sum of all outgoing rates from state $i$.

To construct the matrix $\mathbf{A}$, we first establish a one-to-one mapping between the three-dimensional state space $(n,d,i)$ and a one-dimensional index $k$. We set $n_{max}$, $d_{max}$, and $i_{max}$ as sufficiently large upper bounds for the number of Notch, Delta, and NICD molecules in cell $A$, respectively.  These define the size of our 3D ``box" of possible states.

\textbf{Remark}: These are not strict limits but chosen large enough to technically capture all relevant dynamics.

For a system with sufficiently large values $n_{max}$, $d_{max}$, and $i_{max}$, we now define:

\[
k(n,d,i) = n \cdot (d_{max}+1) \cdot (i_{max}+1) + d \cdot (i_{max}+1) + i
\]
for $0 \leq n \leq n_{max}, \; 0 \leq d \leq d_{max}, \; 0 \leq i \leq i_{max}$.

@@

\begin{align*}
A_{k,k'} = 
\begin{cases}
H^+(i) & \text{if } (n',d',i') = (n+1,d,i) \quad \text{(Notch production)} \\
H^-(i) & \text{if } (n',d',i') = (n,d+1,i) \quad \text{(Delta production)} \\
\gamma(n) & \text{if } (n',d',i') = (n-1,d,i) \quad \text{(Notch decay)} \\
\gamma(d) & \text{if } (n',d',i') = (n,d-1,i) \quad \text{(Delta decay)} \\
\gamma_I(i) & \text{if } (n',d',i') = (n,d,i-1) \quad \text{(NICD decay)} \\
k_T(n)D_{ext} & \text{if } (n',d',i') = (n-1,d,i+1) \quad \text{(Trans-activation)} \\
-v_k & \text{if } k' = k \quad \text{(Diagonal terms)}
\end{cases}
\end{align*}

Note that $A_{k,k'}$ represents the transition rate from state $k$ to state $k'$. The diagonal elements $A_{k,k}$ are the negative sum of all outgoing rates from state $k$:

\[
A_{k,k} = -v_k =-[H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext}]
\]

where $(n,d,i)$ is the state corresponding to index $k$.\\
\textbf{Recall}: We assume that reaction times follow a Poisson point process.

@@

To illustrate, we consider a minimal system with $n_{max} = d_{max} = i_{max} = 1$. This gives us 8 possible states:
$(0,0,0)$, $(0,0,1)$, $(0,1,0)$, $(0,1,1)$, $(1,0,0)$, $(1,0,1)$, $(1,1,0)$, $(1,1,1).$

@@

\small \begin{align*}
\mathbf{A} = 
\begin{pmatrix}
-v_{000} & \alpha_{000,001} & \alpha_{000,010} & \alpha_{000,011} & \alpha_{000,100} & \alpha_{000,101} & \alpha_{000,110} & \alpha_{000,111} \\
\alpha_{001,000} & -v_{001} & \alpha_{001,010} & \alpha_{001,011} & \alpha_{001,100} & \alpha_{001,101} & \alpha_{001,110} & \alpha_{001,111} \\
\alpha_{010,000} & \alpha_{010,001} & -v_{010} & \alpha_{010,011} & \alpha_{010,100} & \alpha_{010,101} & \alpha_{010,110} & \alpha_{010,111} \\
\alpha_{011,000} & \alpha_{011,001} & \alpha_{011,010} & -v_{011} & \alpha_{011,100} & \alpha_{011,101} & \alpha_{011,110} & \alpha_{011,111} \\
\alpha_{100,000} & \alpha_{100,001} & \alpha_{100,010} & \alpha_{100,011} & -v_{100} & \alpha_{100,101} & \alpha_{100,110} & \alpha_{100,111} \\
\alpha_{101,000} & \alpha_{101,001} & \alpha_{101,010} & \alpha_{101,011} & \alpha_{101,100} & -v_{101} & \alpha_{101,110} & \alpha_{101,111} \\
\alpha_{110,000} & \alpha_{110,001} & \alpha_{110,010} & \alpha_{110,011} & \alpha_{110,100} & \alpha_{110,101} & -v_{110} & \alpha_{110,111} \\
\alpha_{111,000} & \alpha_{111,001} & \alpha_{111,010} & \alpha_{111,011} & \alpha_{111,100} & \alpha_{111,101} & \alpha_{111,110} & -v_{111}
\end{pmatrix}
\end{align*}

where:
\begin{align*}
\alpha_{ndi,n'd'i'} = 
\begin{cases}
H^+(i) + H^-(i) + \gamma n + \gamma d + \gamma_I i + k_T n D_{ext} & \text{if } (n,d,i) \rightarrow (n',d',i') \\
0 & \text{otherwise}
\end{cases}
\end{align*}

\textbf{Remark}:\\
Let us consider:

\begin{itemize}

    \item A $k$-cell system with $P(n_1,d_1,i_1,n_2,d_2,i_2,\dots,n_k,d_k,i_k,t+\Delta t)$.\\
    For a one-cell system with $P(n,d,i,t+\Delta t)$ (as shown above), even this simplest case leads to a highly complex system. Extending to $k$ cells leads to a rapid growth in the dimensionality of the state space, making the system increasingly intractable.

    \item To reduce this complexity, we introduce $D_{\text{ext}}$, representing the average number of Delta molecules in the neighbourhood of cell $A$. The exact value of $D_{\text{ext}}$ can vary depending on the choice of domain:
    \begin{itemize}
        \item For two cells, $D_{\text{ext}}$ is the number of Delta molecules in the other cell (cell $B$).
        \item In a linear chain, $D_{\text{ext}}$ is the average over two neighbours.
        \item In a hexagonal lattice, $D_{\text{ext}}$ is the average over six neighbours.
    \end{itemize}
\end{itemize}

Altogether, the complexity of the Kolmogorov forward equation, even for a single cell, motivates the use of numerical methods to model multi-cellular interactions.

@@

In this section, we demonstrate that our stochastic model converges in expectation to the following system of ODEs, adapted from Boareto et al. (2015):
$$
\begin{aligned}
  \frac{dN}{dt} &= \frac{n_{m}I^2}{n_{0}^2 + I^2} - k_{T}ND_{ext} - \gamma N \\[5pt]
  \frac{dD}{dt} &= \frac{d_{m}d_{0}^2}{d_{0}^2 + I^2} - k_{T}DN_{ext} - \gamma D \\[5pt]
  \frac{dI}{dt} &= k_{T}ND_{ext} - \gamma_{I}I
\end{aligned}
$$

This system of ODEs represents the deterministic approximation of the underlying stochastic process, valid when molecular counts are sufficiently large that fluctuations are small compared to mean values.

@@

We define the following expected values:
\begin{itemize}
  \item $\langle n \rangle = \sum_{n,d,i} n \cdot P(n,d,i,t)$ = expected number of Notch receptors
  \item $\langle d \rangle = \sum_{n,d,i} d \cdot P(n,d,i,t)$ = expected number of Delta ligands
  \item $\langle i \rangle = \sum_{n,d,i} i \cdot P(n,d,i,t)$ = expected number of NICD molecules
\end{itemize}

\subsection*{Key Assumptions}

Throughout this derivation, we make the following assumptions:

\begin{enumerate}
  \item The standard deviation of the molecular counts is small compared to their mean values. Specifically:
  \begin{itemize}
    \item $\sqrt{\text{Var}(N)} \ll \langle n \rangle$
    \item $\sqrt{\text{Var}(D)} \ll \langle d \rangle$
    \item $\sqrt{\text{Var}(I)} \ll \langle i \rangle$
  \end{itemize}

  \item As a consequence of assumption 1, we can approximate:
  \begin{itemize}
    \item $\langle n^2 \rangle \approx \langle n \rangle^2$
    \item $\langle d^2 \rangle \approx \langle d \rangle^2$
    \item $\langle i^2 \rangle \approx \langle i \rangle^2$
  \end{itemize}

  \item The molecular counts are approximately independent, allowing us to factorize expected values:
  \begin{itemize}
    \item $\langle n \cdot d \rangle \approx \langle n \rangle \cdot \langle d \rangle$
    \item $\langle n \cdot i \rangle \approx \langle n \rangle \cdot \langle i \rangle$
    \item $\langle d \cdot i \rangle \approx \langle d \rangle \cdot \langle i \rangle$
  \end{itemize}

  \item For functions $H^+$ and $H^-$, we approximate:
  \begin{itemize}
    \item $\langle H^+(i) \rangle \approx H^+(\langle i \rangle)$
    \item $\langle H^-(i) \rangle \approx H^-(\langle i \rangle)$
    \item $\langle n \cdot H^+(i) \rangle \approx \langle n \rangle \cdot H^+(\langle i \rangle)$
    \item $\langle d \cdot H^-(i) \rangle \approx \langle d \rangle \cdot H^-(\langle i \rangle)$
  \end{itemize}
\end{enumerate}

\subsection*{Derivation of ODE for Expected Notch ($\langle n \rangle$)}

Starting with the time derivative of the expected value of Notch:
$$\frac{d\langle n \rangle}{dt} = \frac{d}{dt}\sum_{n,d,i} n \cdot P(n,d,i,t) = \sum_{n,d,i} n \cdot \frac{\partial P(n,d,i,t)}{\partial t}$$
Substituting the Kolmogorov forward equation and analysing each term:

\subsubsection*{Notch Production Term}
Using index shifting with $n' = n-1$ (so $n = n'+1$):
\begin{align*}
\sum_{n,d,i} n \cdot P(n-1,d,i,t) \cdot H^+(i) &= \sum_{n',d,i} (n'+1) \cdot P(n',d,i,t) \cdot H^+(i) \\
&= \sum_{n',d,i} n' \cdot P(n',d,i,t) \cdot H^+(i) + \sum_{n',d,i} P(n',d,i,t) \cdot H^+(i) \\
&= \langle n \cdot H^+(i) \rangle + \langle H^+(i) \rangle \\
&\approx \langle n \rangle \cdot H^+(\langle i \rangle) + H^+(\langle i \rangle) \\
&= H^+(\langle i \rangle) \cdot (1 + \langle n \rangle)
\end{align*}

\subsubsection*{Delta Production Term}
This term does not affect $\langle n \rangle$ directly:
\begin{align*}
\sum_{n,d,i} n \cdot P(n,d-1,i,t) \cdot H^-(i) &= \langle n \cdot H^-(i) \rangle
\approx \langle n \rangle \cdot H^-(\langle i \rangle)
\end{align*}

\subsubsection*{Notch Decay Term}
Using index shifting with $n' = n+1$ (so $n = n'-1$):
\begin{align*}
\sum_{n,d,i} n \cdot P(n+1,d,i,t) \cdot \gamma(n+1) &= \sum_{n',d,i} (n'-1) \cdot P(n',d,i,t) \cdot \gamma n' \\
&= \gamma \sum_{n',d,i} (n'-1)n' \cdot P(n',d,i,t) \\
&= \gamma \sum_{n',d,i} n'^2 \cdot P(n',d,i,t) - \gamma \sum_{n',d,i} n' \cdot P(n',d,i,t) \\
&= \gamma(\langle n^2 \rangle - \langle n \rangle) \\
&\approx \gamma(\langle n \rangle^2 - \langle n \rangle) \\
&= \gamma \langle n \rangle (\langle n \rangle - 1)
\end{align*}

\subsubsection*{Delta Decay Term}
\begin{align*}
\sum_{n,d,i} n \cdot P(n,d+1,i,t) \cdot \gamma(d+1) &= \gamma \sum_{n,d',i} n \cdot P(n,d',i,t) \cdot d' = \gamma \langle n \cdot d \rangle \approx \gamma \langle n \rangle \langle d \rangle
\end{align*}
